On-the-fly Algorithms for GKAT Equivalences

### 2025-01-24 main.tex
on the fly bisimulation

In this paper, we will make heavy use of coalgebraic theory, thus we will recall useful notions and theorems in universal coalgebra.
Given a functor \(F\) on the category of set and functions, a \emph{coalgebra over \(F\)} or \emph{\(F\)-coalgebra} consists of a set \(S\) and a function \(σ_S: S → F(S)\).
We typically call elements in \(S\) the \emph{states} of the coalgebra, and \(σ_S(s)\) the \emph{dynamic} of state \(s\).
We will sometimes use the states \(S\) to denote the coalgebra when no ambiguity can arise.

A homomorphism between two \(F\)-coalgebra \(S\) and \(U\) is a map \(h: S → U\) that preserves the function \(σ\); diagrammatically, the following diagram commutes:
\[
    \begin{tikzcd}
        S \ar{r}{h} \ar[swap]{d}{σ_S} & U \ar{d}{σ_U} \\  
        F(S) \ar{r}{F(h)} & F(U)
    \end{tikzcd}    
\]

When we can restrict the homomorphism map into an inclusion map \(i: S' → S\) for \(S' ⊆ S\) then we say that \(S'\) is a \emph{sub-coalgebra} of \(S\), denoted as \(S' ⊑ S\). Specifically, the following diagram commutes when \(S' ⊑ S\):
\[
    \begin{tikzcd}
        S' \ar[hook]{r}{i} \ar[swap]{d}{σ_{S'}} & S \ar{d}{σ_S} \\  
        F(S') \ar[hook]{r}{F(i)} & F(S)
    \end{tikzcd}    
\]
In fact, the function \(σ_{S'}\) is uniquely determined by the states \(S'\)~\cite[Proposition 6.1]{rutten_UniversalCoalgebraTheory_2000}.
Sub-coalgebras are also preserved under homomorphic images and pre-images: 
\begin{theoremEnd}{lemma}[Theorem 6.3~\cite{rutten_UniversalCoalgebraTheory_2000}]\label{thm:hom-(pre)img-preserve-sub-coalg}
    given a homomorphism \(h: S → U\), and sub-coalgebras \(S' ⊑ S\) and \(U' ⊑ U\), then 
    \[h(S') ⊑ U \text{ and } h^{-1}(U') ⊑ S.\]
\end{theoremEnd}

One particularly important sub-coalgebra of a coalgebra \(S\) is the least sub-coalgebra that contains a state \(s\). 
We will denote this sub-coalgebra as \(⟨s⟩_{S}\), and call it \emph{principle sub-coalgebra} generated by \(s\). 
We sometimes omit the subscript \(S\) when it can be inferred from context or irrelevant.
Intuitively, we usually think of principle sub-coalgebra \(⟨s⟩_S\) as the sub-coalgebra that is formed by all the ``reachable state'' from \(s\).
This coalgebraic characterization of reachable state can allow us to avoid induction on the length of a path from \(s\) to a state in \(⟨s⟩_S\).

For all coalgebra \(S\) and a state \(s ∈ S\), principle sub-coalgebra \(⟨s⟩_S\) always exists and is unique, because sub-coalgebra of any coalgebra forms a complete lattice~\cite[theorem 6.4]{rutten_UniversalCoalgebraTheory_2000}; thus taking the meet of all the sub-coalgebra that contains \(s\) will yield \(⟨s⟩_S\).

@@

Given two \(F\)-coalgebra \(S\) and \(U\), the \emph{behavioral equivalence} between states in \(S\) and \(U\) can be computed by a notion called \emph{bisimulation}.
A relation \({∼} ⊆ S × U\) is called a \emph{bisimulation relation} if it forms an \(F\)-coalgebra: \[σ_{∼}: {∼} → F(∼),\] 
and its projections \(π₁: {∼} → S\) and \(π₂: {∼} → U\) are both homomorphisms:
\[
    \begin{tikzcd}[column sep=1.25cm]
        S \ar[swap]{d}{σ_S}
            & ∼ \ar[swap]{l}{π₁} \ar{r}{π₂} \ar{d}{σ_{∼}} 
            & U \ar{d}{σ_T}\\  
        F(S) & F(∼) \ar{l}{F(π₁)} \ar[swap]{r}{F(π₂)} & F(T)
    \end{tikzcd}
\]
In a special case, when there exists a homomorphism \(h: S → U\), we can simply pick \({∼} ⊆ S × U\) to be \(\{(s, h(s)) ∣ s ∈ S\}\), which gives us a bisimulation with the lift \(σ_{∼} ≜ σ_S × σ_U\).

When given a homomorphism \(h: S → U\), we can construct a bisimulation \({∼} ⊆ S × U\) where it relates all the states \(s ∈ S\) with \(h(s) ∈ U\). Each component of the pair will transition individually via their respective transition function \(δ_∼ ≜ δ_S × δ_U\) where \(×\) is the bifunctorial lift of product.
\begin{theoremEnd}{corollary}\label{thm:hom-preserves-semantics}
    Given a homomorphism \(h: S → U\) between two \(F\)-coalgebra \(S, U\), then for all states \(s ∈ S\), \(\mathrm{beh}_S(s) = \mathrm{beh}_U(h(s)).\)
\end{theoremEnd}

@@

Guarded Kleene Algebra with Tests, or GKAT~\cite{smolka_GuardedKleeneAlgebra_2020}, is a deterministic fragment of Kleene Algebra with Tests. 
The syntax of GKAT over a set of primitive actions \(K\) and a set of primitive tests \(T\) can be defined in two sorts, boolean expressions \(\BExp\) and GKAT expressions \(\Exp\):
\begin{align*}
    a, b, c ∈ \BExp 
        & ≜ 1 ∣ 0 ∣ t ∈ T ∣ b ∧ c ∣ b ∨ c ∣ \overline{b} \\  
    e, f ∈ \Exp 
        & ≜ p ∈ K ∣ b ∈ \BExp ∣ e +_b f ∣ e ; f ∣ e^{(b)} 
\end{align*}
where \(e +_b f\) is the if statement with condition \(b\), \(e;f\) is the sequencing of expression \(e\) and \(f\), and \(e^{(b)}\) is the while loop with body \(e\) and condition \(b\).
We use notation like \(b ≤ c\), \(b ≡ c\), and \(b ≢ c\) for the usual order, equivalence, and inequivalence in Boolean Algebra.
A GKAT expression can be unfolded into a KAT expression in the usual manner~\cite{kozen_KleeneAlgebraTests_1997c}:
\begin{align*}
    e +_b f & ≜ b; e + \overline{b}; f &
    e^{(b)} & ≜ (b; e)^*; \overline{b}.
\end{align*}
Then the semantics of each expression \(⟦e⟧\) can be computed by the semantics of Kleene Algebra with Tests~\cite{kozen_KleeneAlgebraTests_1997c}.
An important construct in the semantics is \emph{atoms}, which are conjunctions of all the primitive tests either in its positive or negative form: for \(T ≜ \{t₁, t₂, …, tₙ\}\)
\[\At_T ≜ \{t₁' ∧ t₂' ∧ ⋯ ∧ tₙ' ∣ tᵢ' ∈ \{tᵢ, \overline{tᵢ}\}\}.\]
Following the conventional notation, we denote an atom using \(α, β\); and we sometimes omit the subscript \(T\) when no confusion can arise.
Alternatively, atoms can also be thought of as truth assignments to each primitive test, indicating which primitive tests are satisfied in the current program state; and \(α ≤ b\) if and only if the truth assignment represented by \(α\) satisfies \(b\).

@@

Formally, GKAT coalgebras over primitive actions \(K\) and primitive tests \(T\) are coalgebras over the following functor:
\[G(S) ≜ (\{\accept, \reject\} + S × K)^{\At_T}.\] 
Intuitively, given a state \(s ∈ S\) and an atom \(α ∈ \At\), \(δ(s, α)\) will deterministically execute one of the following: reject \(α\) when \(δ(s, α) = \reject\); accept \(α\) when \(δ(s, α) = \accept\); or transition to a state \(s' ∈ S\) and execute action \(p ∈ K\) when \(δ(s, α) = (s', p)\).

@@

Our algorithm is lazy in the dead states detection i.e. the dead states are only checked when a mismatch requires it; even then, our dead state detection algorithm will only check the necessary states to compute the liveness of the states causing the mismatch.
This not only avoids unnecessarily checking the liveness of a state, but also enables on-the-fly generation of the coalgebra similarly to using derivatives where we can remove irrelevant states from memory after they have been explored.

@@

Traditionally, live and dead states are defined by whether they can reach an accepting state~\cite{smolka_GuardedKleeneAlgebra_2020}.
\begin{definition}[Liveness]\label{def:liveness-of-states}
    A state \(s\) is \emph{accepting} if there exists an atom \(α ∈ \At\) s.t. \(δ(s, α) = \accept\).
    A state \(s₀\) is \emph{live} if there exists an accepting state \(sₙ\) s.t. there is a liveness chain:
    \[s₀ \transvia{α₁ ∣ p₁} s₁ \transvia{α₂ ∣ p₂} s₂ ⋯ \transvia{αₙ ∣ pₙ} sₙ.\]
    A state is \emph{dead} if it is not live.
\end{definition}
However, besides directly working with this explicit definition, we can also give a coalgebraic characterization of these properties.
Specifically, by induction on the length of the path from one state to the next, we can show that the principle sub-coalgebra \(⟨s⟩_S\) exactly contains all the reachable states of \(s\) in the GKAT coalgebra \(S\).
Thus we obtain the following coalgebraic characterization of liveness:
\begin{theoremEnd}{theorem}\label{thm:liveness-principle-sub-coalg}
    A state \(s\) is \emph{live} if there exists an accepting state \(s' ∈ ⟨s⟩\); and a state \(s\) is \emph{dead} if there is no accepting state in \(⟨s⟩\).
\end{theoremEnd}
This alternative liveness definition can help us prove important theorems regarding reachability and liveness without explicitly performing induction on the liveness chain. 
We show the following theorems as examples:
\begin{theoremEnd}{lemma}\label{thm:dead-iff-all-reachable-dead}
    A state \(s\) is dead if and only if all elements in \(⟨s⟩\) are dead.
\end{theoremEnd}
\begin{proofEnd}
    \(⟸\) direction is true, because \(s ∈ ⟨s⟩\): if all \(⟨s⟩\) is dead, then \(s\) is dead. 
    \(⟹\) direction can be proven as follows.
    Take any \(s' ∈ ⟨s⟩\), then \(⟨s'⟩ ⊑ ⟨s⟩\) because \(s'\) is the minimal sub-coalgebra that contains \(s'\). 
    Since there is no accepting state in \(⟨s⟩\), thus there cannot be any accepting state in \(⟨s'⟩\), hence \(s'\) is also dead.
\end{proofEnd}

\begin{theoremEnd}{theorem}[homomorphism perserves liveness]\label{thm:hom-preserve-liveness}
    Given a homomorphism \(h: S → U\) and a state \(s ∈ S\):
    \[\text{\(s\) is live} ⟺ \text{\(h(s)\) is live}\]
\end{theoremEnd}

@@

Because of the functoriality, we can show that two states that are infinite-trace equivalent implies these two states are finite-trace equivalent.
This gives us more tool in proving semantic equivalence between two states in GKAT coalgebras: proving bisimulation of these two states in the \emph{non-normalized} GKAT coalgebra can also obtain semantic equivalence for two states.

@@

The original algorithm for deciding GKAT equivalences~\cite{smolka_GuardedKleeneAlgebra_2020} requires the entire coalgebra to be known prior to the execution of the bisimulation algorithm; specifically, it is necessary to iterate through the entire coalgebra in order to identify the liveness of every single states and perform the normalization operation.
This limitation poses challenges to the design of an efficient on-the-fly algorithm for GKAT.
To make the equivalence-checking procedure scalable, we propose to greedily run the bisimulation algorithm and only invoke liveness detection when a discrepancy is found. 
For example, when deciding the trace equivalence between \(s ∈ S\) and \(u ∈ U\), if \(δ_S(s, α) = (s', p)\) and \(δ_U(u, α) = (u', p)\), we will proceed to recurse on \(s'\) and \(u'\), without checking the liveness of \(s'\) and \(u'\).
However, if \(s\) rejects \(α\), instead of transitioning to \(s'\), we will check if \(u'\) is dead.

This approach offers several benefit: first, because the liveness detection is only called when necessary, it is unlikely to iterate through the entire coalgebra; second, the algorithm can short-circuit on a counter-example, allowing termination before all the states are explored; finally, the algorithm supports on-the-fly constructions like derivatives as the coalgebra is no longer required to be fully constructed.

The correctness argument of this on-the-fly algorithm is centered around the following coalgebraic structure, which we named ``greedy bisimulation''.

@@

Although the proof of~\Cref{thm:dead-construction-greedy-bisim} basically proceeds by unfolding the definition, \(∼_d\) turns out to be an important structure. 
Specifically, \(∼_d\) serves a connection between \(≈\) and its underlying bisimulation \(∼\), and the order \({∼} ⊆ {≈} ⊑ {~_d}\) will reveal many structural properties of \(≈\).
We show the following corollary to demonstrate the usefulness of \(∼_d\).

\begin{theoremEnd}{corollary}\label{thm:greedy-bisim-exists-unique}
    For any bisimulation between two normalized GKAT coalgebras \({∼} ⊆ \norm(S) × \norm(U)\), a greedy bisimulation \(≈\) with \(∼\) as underlying bisimulation exists and is unique.
\end{theoremEnd}

@@

\begin{theoremEnd}{corollary}\label{thm:greedy-bisim-dead-or-bisim}
    For every pair of states \((s, u)\) in a greedy bisimulation \({≈} ⊆ S × U\), either \((s, u)\) is in the underlying bisimulation \(∼\), or both \(s\) and \(u\) are dead in \(S\) and \(U\).
\end{theoremEnd}

\begin{proofEnd}
    Because \(∼_d\) is a GKAT coalgebra with transition \(δ_≈\) and also contains all the states in \(∼\), therefore \({≈} ⊑ {∼_d}\).
    Thus, for all \((s, u) ∈ {≈}\) but not in \(∼\), \(s\) and \(u\) must be dead in \(S\) and \(U\).
\end{proofEnd}

\begin{theoremEnd}{corollary}[Perservation Of Liveness]\label{thm:greedy-bisim-perserve-liveness}
    Given a pair of states in a greedy bisimulation \(s ≈ u\) where \({≈} ⊆ S × U\), then the following are equivalent
    \begin{itemize}
        \item \((s, u)\) is live (or dead) in \({≈}\);
        \item \(s\) is live (or dead) in \(S\);  
        \item \(u\) is live (or dead) in \(U\).
    \end{itemize}
    Furthermore, if \((s, u)\) is in the underlying bisimulation \(∼\) of \(≈\), then all of above are also equivalent to \((s, u)\) is live (or dead) in \(∼\).
\end{theoremEnd}

\begin{proofEnd}
    Let \(∼\) by the underlying bisimulation of \(≈\), consider the \(∼_d\) in~\Cref{thm:dead-construction-greedy-bisim}, then we know that \({≈} ⊑ {∼_d}\).

    Because \(\norm(∼_d)\) is a bisimulation between \(\norm(S)\) and \(\norm(U)\), then
    \begin{align*}
        & \text{\(s\) is live in \(S\)} \\
        \text{ iff } & \text{\(s\) is live in \(\norm(S)\)} 
            & \text{\Cref{thm:norm-perserve-liveness}}\\  
        \text{ iff } & \text{\((s, u)\) is live in \(\norm(∼_d)\)} 
            & \text{\Cref{thm:bisim-preserve-liveness}}\\
        \text{ iff } & \text{\((s, u)\) is live in \(∼_d\)} 
            & \text{\Cref{thm:norm-perserve-liveness}}\\
        \text{ iff } & \text{\((s, u)\) is live in \(≈\)} 
            & \text{\Cref{thm:sub-coalg-preserve-liveness}}
    \end{align*}
\end{proofEnd}

\begin{theoremEnd}{corollary}\label{thm:greedy-bisim-iff-norm-bisim}
    Given two GKAT coalgebras \(S\) and \(U\), and a relational GKAT coalgebra \({≈} ⊆ S × U\) with the transition \(δ_≈\) as in~\cref{def:greedy-bisim}.
    \(≈\) is a greedy bisimulation if and only if \(\norm(≈)\) is a bisimulation between \(\norm(S)\) and \(\norm(U)\)
\end{theoremEnd}

@@

\begin{theoremEnd}{corollary}\label{thm:max-bisim-is-max-greedy-bisim}
    Given two GKAT coalgebra \(S\) and \(U\), carrier of the maximal bisimulation \({≡} ⊆ \norm(S) × \norm(U)\) is also the carrier of maximal greedy bisimulation, which we will denote as \(≊ ⊆ S × U\).
    We will abuse the notation \(≡\) to represent both the maximal bisimulation and maximal greedy bisimulation.
\end{theoremEnd}

\begin{proofEnd}
    by definition of greedy bisimulation, the carrier of \({≡}\) is a subset of the carrier of \({≊}\). 
    Then because of~\Cref{thm:greedy-bisim-iff-norm-bisim}, \(\norm(≊)\) is a bisimulation between \(\norm(S)\) and \(\norm(U)\), which implies \(\norm(≡') ⊑ {≡}\).
    Finally because \(\norm\) do not change the carrier set, therefore the carrier of \(≡\) and \(≊\) necessarily coincide.

    Then we will need to show that \(≡\) with the \(δ_≈\) defined in~\Cref{def:greedy-bisim} is the maximal greedy bisimulation.
    Take an arbitrary greedy bisimulation \({≈} ⊆ S × U\) with underlying bisimulation \(∼\), and a pair of states \(s ≈ u\), we will need to show \(s ≡ u\).
    By~\cref{thm:greedy-bisim-dead-or-bisim}, we can case analysis on \(s\) and \(u\): if \(s ∼ u\), then \(s ≡ u\) because \(≡\) is the maximal bisimulation; and if \(s\) and \(u\) are both dead, then by~\cref{thm:bisim-between-dead}, \(s ≡ u\).
\end{proofEnd}

\Cref{thm:max-bisim-is-max-greedy-bisim} allows us to take the maximal bisimulation as a greedy bisimulation, aiding several proofs.

@@

\begin{theoremEnd}{theorem}[Correctness]
    Given two states in GKAT coalgebras \(s ∈ S\) and \(u ∈ ,U\), there exists a greedy bisimulation \({≈} ⊆ S × U\) s.t. \(s ≈ u\), if and only if \(⟦s⟧_S = ⟦u⟧_U\).
\end{theoremEnd}

@@

Before we introduce the optimized algorithm, we will first briefly sketch the liveness-check algorithm. It uses depth-first search to check whether there exists any accepting states in the reachable states of \(s ∈ S\): if there exists an accepting state in \(⟨s⟩_S\), then the algorithm terminates immediately, and returns that \(s\) is live, otherwise it would return all the states in \(⟨s⟩\), and by~\cref{thm:dead-iff-all-reachable-dead}, all the states in \(⟨s⟩\) are dead. Note that we use depth-first search here for its simplicity, other search algorithms will also be sound.

>>>>

In this paper, we will make heavy use of coalgebraic theory, thus we will recall useful notions and theorems in universal coalgebra.
Given a functor \(F\) on the category of set and functions, a \emph{coalgebra over \(F\)} or \emph{\(F\)-coalgebra} consists of a set \(S\) and a function \(σ_S: S → F(S)\).
We typically call elements in \(S\) the \emph{states} of the coalgebra, and \(σ_S(s)\) the \emph{dynamic} of state \(s\).
We will sometimes overload \(S\) to denote both the state set and the coalgebra, when no ambiguity can arise.

A homomorphism between two \(F\)-coalgebra \(S\) and \(U\) is a map \(h: S → U\) that preserves the function \(σ\):
\[
    \begin{tikzcd}
        S \ar{r}{h} \ar[swap]{d}{σ_S} & U \ar{d}{σ_U} \\  
        F(S) \ar{r}{F(h)} & F(U)
    \end{tikzcd}    
\]

When we can restrict the homomorphism map into an inclusion map \(i: S' → S\) for \(S' ⊆ S\) then we say that \(S'\) is a \emph{sub-coalgebra} of \(S\), denoted as \(S' ⊑ S\).
In fact, the function \(σ_{S'}\) is uniquely determined by the states \(S'\)~\cite[Proposition 6.1]{rutten_UniversalCoalgebraTheory_2000}.
Sub-coalgebras are also preserved under homomorphic images and pre-images: 
\begin{theoremEnd}{lemma}[Theorem 6.3~\cite{rutten_UniversalCoalgebraTheory_2000}]\label{thm:hom-(pre)img-preserve-sub-coalg}
    given a homomorphism \(h: S → U\), and sub-coalgebras \(S' ⊑ S\) and \(U' ⊑ U\), then 
    \[h(S') ⊑ U \text{ and } h^{-1}(U') ⊑ S.\]
\end{theoremEnd}

One particularly important sub-coalgebra of a coalgebra \(S\) is the least sub-coalgebra that contains a state \(s\). 
We will denote this sub-coalgebra as \(⟨s⟩_{S}\), and call it \emph{principle sub-coalgebra} generated by \(s\). 
We sometimes omit the subscript \(S\) when it can be inferred from context or irrelevant.
Intuitively, we usually think of principle sub-coalgebra \(⟨s⟩_S\) as the sub-coalgebra that is formed by all the ``reachable states'' from \(s\).
For all coalgebra \(S\) and a state \(s ∈ S\), principle sub-coalgebra \(⟨s⟩_S\) always exists and is unique, because sub-coalgebra of any coalgebra forms a complete lattice~\cite[theorem 6.4]{rutten_UniversalCoalgebraTheory_2000}; thus taking the meet of all the sub-coalgebra that contains \(s\) will yield \(⟨s⟩_S\).

@@

Given two \(F\)-coalgebra \(S\) and \(U\), the \emph{behavioral equivalence} between states in \(S\) and \(U\) can be computed by a notion called \emph{bisimulation}.
A relation \({∼} ⊆ S × U\) is called a \emph{bisimulation relation} if it forms an \(F\)-coalgebra: \[σ_{∼}: {∼} → F(∼),\] 
and its projections \(π₁: {∼} → S\) and \(π₂: {∼} → U\) are both homomorphisms:
\[
    \begin{tikzcd}[column sep=1.25cm]
        S \ar[swap]{d}{σ_S}
            & ∼ \ar[swap]{l}{π₁} \ar{r}{π₂} \ar{d}{σ_{∼}} 
            & U \ar{d}{σ_T}\\  
        F(S) & F(∼) \ar{l}{F(π₁)} \ar[swap]{r}{F(π₂)} & F(T)
    \end{tikzcd}
\]
We sometimes use \({∼} ⊆ S × U\) to specify that \(∼\) is a bisimulation between \(S\) and \(U\).

When given a homomorphism \(h: S → U\), we can construct a bisimulation \({∼} ⊆ S × U\) where it relates all the states \(s ∈ S\) with \(h(s) ∈ U\). Each component of the pair will transition individually via their respective transition function \(δ_∼ ≜ δ_S × δ_U\).
\begin{theoremEnd}{corollary}\label{thm:hom-preserves-semantics}
    Given a homomorphism \(h: S → U\) between two \(F\)-coalgebra \(S, U\), then for all states \(s ∈ S\), \(\mathrm{beh}_S(s) = \mathrm{beh}_U(h(s)).\)
\end{theoremEnd}

@@

Guarded Kleene Algebra with Tests, or GKAT~\cite{smolka_GuardedKleeneAlgebra_2020}, is a deterministic fragment of Kleene Algebra with Tests (KAT). 
The syntax of GKAT over a set of primitive actions \(K\) and a set of primitive tests \(T\) can be defined in two sorts, boolean expressions \(\BExp\) and GKAT expressions \(\Exp\):
\begin{align*}
    a, b, c ∈ \BExp 
        & ≜ 1 ∣ 0 ∣ t ∈ T ∣ b ∧ c ∣ b ∨ c ∣ \overline{b} \\  
    e, f ∈ \Exp 
        & ≜ p ∈ K ∣ b ∈ \BExp ∣ e +_b f ∣ e ; f ∣ e^{(b)} 
\end{align*}
where \(e +_b f\) is the if statement with condition \(b\), \(e;f\) is the sequencing of expression \(e\) and \(f\), and \(e^{(b)}\) is the while loop with body \(e\) and condition \(b\).
We use notation like \(b ≤ c\), \(b ≡ c\), and \(b ≢ c\) for the usual order, equivalence, and inequivalence in Boolean Algebra.
A GKAT expression can be unfolded into a KAT expression in the usual manner~\cite{kozen_KleeneAlgebraTests_1997c}:
\begin{align*}
    e +_b f & ≜ b; e + \overline{b}; f &
    e^{(b)} & ≜ (b; e)^*; \overline{b}.
\end{align*}
Then the semantics of each expression \(⟦e⟧\) can be computed by the semantics of Kleene Algebra with Tests~\cite{kozen_KleeneAlgebraTests_1997c}.
An important construct in the semantics is \emph{atoms}, which are conjunctions of all the primitive tests either in its positive or negative form: for \(T ≜ \{t₁, t₂, …, tₙ\}\),
\[\At_T ≜ \{t₁' ∧ t₂' ∧ ⋯ ∧ tₙ' ∣ tᵢ' ∈ \{tᵢ, \overline{tᵢ}\}\}.\]
Following the conventional notation, we denote an atom using \(α, β\); and we sometimes omit the subscript \(T\) when no confusion can arise.
Alternatively, atoms can also be thought of as truth assignments to each primitive test, indicating which primitive tests are satisfied in the current program state; and \(α ≤ b\) in Boolean Algebra if and only if the truth assignment represented by \(α\) satisfies \(b\).

@@

Formally, GKAT coalgebras over primitive actions \(K\) and primitive tests \(T\) are coalgebras over the following functor:
\[G(S) ≜ (\{\accept, \reject\} + S × K)^{\At_T}.\] 
Intuitively, given a state \(s ∈ S\) and an atom \(α ∈ \At\), \(δ_S(s, α)\) will deterministically execute one of the following: reject \(α\) when \(δ(s, α) = \reject\), denoted as \(s \transRej{S}{α}\); accept \(α\) when \(δ(s, α) = \accept\), denoted \(s \transAcc{S}{α}\); or transition to a state \(s' ∈ S\) and execute action \(p ∈ K\) when \(δ(s, α) = (s', p)\), denoted \(s \transvia{α ∣ p}_S s'\).

@@

Our algorithm is lazy in the dead states detection i.e. the dead states are only checked when a mismatch requires it; even then, our dead state detection algorithm will only check the necessary states to compute the liveness of the states causing the mismatch.
This not only avoids unnecessarily checking the liveness of a state, but also enables on-the-fly generation of the coalgebra, where we can remove irrelevant states from memory after they have been explored.

@@

Traditionally, live and dead states are defined by whether they can reach an accepting state~\cite{smolka_GuardedKleeneAlgebra_2020}.
\begin{definition}[Liveness]\label{def:liveness-of-states}
    A state \(s\) is \emph{accepting} if there exists an atom \(α ∈ \At\) s.t. \(δ(s, α) = \accept\).
    A state \(s₀\) is \emph{live} if there exists an accepting state \(sₙ\) s.t. there is a liveness chain:
    \[s₀ \transvia{α₁ ∣ p₁} s₁ \transvia{α₂ ∣ p₂} s₂ ⋯ \transvia{αₙ ∣ pₙ} sₙ.\]
    A state is \emph{dead} if it is not live.
\end{definition}
However, besides directly working with this explicit definition, we can also give a coalgebraic characterization of these properties.
Specifically, by induction on the length of the path from one state to the next, we can show that the principle sub-coalgebra \(⟨s⟩_S\) exactly contains all the reachable states of \(s\) in the GKAT coalgebra \(S\).
Thus we obtain the following coalgebraic characterization of liveness:
\begin{theoremEnd}{theorem}\label{thm:liveness-principle-sub-coalg}
    A state \(s\) is \emph{live} if there exists an accepting state \(s' ∈ ⟨s⟩\); and a state \(s\) is \emph{dead} if there is no accepting state in \(⟨s⟩\).
\end{theoremEnd}
This alternative liveness definition can help us prove important theorems regarding reachability and liveness without explicitly performing induction on the liveness chain. 
We show the following theorems as examples:
\begin{theoremEnd}[normal]{lemma}\label{thm:dead-iff-all-reachable-dead}
    A state \(s\) is dead if and only if all elements in \(⟨s⟩\) are dead.
\end{theoremEnd}
\begin{proofEnd}
    \(⟸\) direction is true, because \(s ∈ ⟨s⟩\): if all \(⟨s⟩\) is dead, then \(s\) is dead. 
    \(⟹\) direction can be proven as follows.
    Take any \(s' ∈ ⟨s⟩\), then \(⟨s'⟩ ⊑ ⟨s⟩\) because \(s'\) is the minimal sub-coalgebra that contains \(s'\). 
    Since there is no accepting state in \(⟨s⟩\), thus there cannot be any accepting state in \(⟨s'⟩\), hence \(s'\) is also dead.
\end{proofEnd}

\begin{theoremEnd}[normal]{theorem}[homomorphism perserves liveness]\label{thm:hom-preserve-liveness}
    Given a homomorphism \(h: S → U\) and a state \(s ∈ S\):
    \[\text{\(s\) is live} ⟺ \text{\(h(s)\) is live}\]
\end{theoremEnd}

@@

Because of the functoriality, we can show that two states that are infinite-trace equivalent implies these two states are finite-trace equivalent.
This gives us more tool in proving semantic equivalence between two states in GKAT coalgebras: proving bisimulation of these two states in the \emph{non-normalized} GKAT coalgebra can also obtain finite-trance equivalence for the two states.

@@

The original algorithm for deciding GKAT equivalences~\cite{smolka_GuardedKleeneAlgebra_2020} requires the entire coalgebra to be known prior to the execution of the bisimulation algorithm; specifically, it is necessary to iterate through the entire coalgebra in order to identify the liveness of every single states and perform the normalization operation.
This limitation poses challenges to the design of an efficient on-the-fly algorithm for GKAT.
To make the equivalence-checking procedure scalable, we propose to greedily run the bisimulation algorithm and only invoke liveness detection when a discrepancy is found. 
For example, when constructing the bisimulation between \(s ∈ S\) and \(u ∈ U\), if \(s \transvia{α ∣ p}_S s'\) and \(u \transvia{α ∣ p}_U u'\), we will proceed to recurse on \(s'\) and \(u'\), without checking the liveness of \(s'\) and \(u'\).
Whereas, if \(s \transRej{S}{α}\) but \(u \transvia{α ∣ p}_U u'\), then the transition of \(s\) and \(u\) disagree, requiring \(u'\) to be dead.

This approach offers several benefit: first, the liveness detection is only called when necessary, thus unlikely to iterate through the entire coalgebra; second, the algorithm can short-circuit on a counter-example, allowing termination before all the states are explored; finally, the algorithm supports on-the-fly coalgebra constructions like derivatives, as the coalgebra is no longer required to be fully constructed before bisimulation.

The intuition of this on-the-fly algorithm can be expressed as the following coalgebraic structure, which we named ``greedy bisimulation''.

@@

Although \Cref{thm:dead-construction-greedy-bisim} can be proved by unfolding the definition, \(∼_d\) turns out to be an important structure. 
Specifically, \(∼_d\) serves a connection between \(≈\) and its underlying bisimulation \(∼\), and the order \({∼} ⊆ {≈} ⊑ {∼_d}\) will reveal many structural properties of \(≈\).

\begin{theoremEnd}[normal]{corollary}\label{thm:greedy-bisim-exists-unique}
    For any bisimulation between two normalized GKAT coalgebras \({∼} ⊆ \norm(S) × \norm(U)\), a greedy bisimulation \(≈\) with \(∼\) as underlying bisimulation exists and is unique.
\end{theoremEnd}

@@

\begin{theoremEnd}[normal]{corollary}\label{thm:greedy-bisim-dead-or-bisim}
    For every pair of states \((s, u)\) in a greedy bisimulation \({≈} ⊆ S × U\), either \((s, u)\) is in the underlying bisimulation \(∼\), or both \(s\) and \(u\) are dead in \(S\) and \(U\).
\end{theoremEnd}

\begin{proofEnd}
    Because \({≈} ⊑ {∼_d}\), and \(∼_d\) only contains pairs in \(∼\) or pairs of dead-state, then so is \(≈\).
\end{proofEnd}

\begin{theoremEnd}[normal]{corollary}[Perservation Of Liveness]\label{thm:greedy-bisim-perserve-liveness}
    Given a pair of states in a greedy bisimulation \(s ≈ u\) where \({≈} ⊆ S × U\), then the following are equivalent
    \begin{itemize}
        \item \((s, u)\) is live (or dead) in \({≈}\);
        \item \(s\) is live (or dead) in \(S\);  
        \item \(u\) is live (or dead) in \(U\).
    \end{itemize}
    Furthermore, if \((s, u)\) is in the underlying bisimulation \(∼\) of \(≈\), then all of above are also equivalent to \((s, u)\) is live (or dead) in \(∼\).
\end{theoremEnd}

\begin{proofEnd}
    Let \(∼\) by the underlying bisimulation of \(≈\), consider the \(∼_d\) in~\Cref{thm:dead-construction-greedy-bisim}.
    Because \({≈} ⊑ {∼_d}\) and \(\norm(∼_d) ⊆ \norm(S) × \norm(U)\) is a bisimulation,
    \begin{align*}
        & \text{\(s\) is live in \(S\)} \\
        \text{ iff } & \text{\(s\) is live in \(\norm(S)\)} 
            & \text{\Cref{thm:norm-perserve-liveness}}\\  
        \text{ iff } & \text{\((s, u)\) is live in \(\norm(∼_d)\)} 
            & \text{\Cref{thm:bisim-preserve-liveness}}\\
        \text{ iff } & \text{\((s, u)\) is live in \(∼_d\)} 
            & \text{\Cref{thm:norm-perserve-liveness}}\\
        \text{ iff } & \text{\((s, u)\) is live in \(≈\)} 
            & \text{\Cref{thm:sub-coalg-preserve-liveness}}
    \end{align*}
\end{proofEnd}

\begin{theoremEnd}[normal]{corollary}\label{thm:greedy-bisim-iff-norm-bisim}
    Given two GKAT coalgebras \(S\) and \(U\), and a relational GKAT coalgebra \({≈} ⊆ S × U\) with the transition \(δ_≈\) as in~\cref{def:greedy-bisim}.
    \(≈\) is a greedy bisimulation if and only if \(\norm(≈)\) is a bisimulation between \(\norm(S)\) and \(\norm(U)\)
\end{theoremEnd}

@@

Finally, with all the structural properties in place, we can derive that the maximal greedy bisimulation actually corresponds to maximal bisimulation.

\begin{theoremEnd}{theorem}\label{thm:max-bisim-is-max-greedy-bisim}
    Given two GKAT coalgebra \(S\) and \(U\), carrier of the maximal bisimulation \({≡} ⊆ \norm(S) × \norm(U)\) is also the carrier of maximal greedy bisimulation, which we will denote as \({≊} ⊆ S × U\).
\end{theoremEnd}

\begin{proofEnd}
    by definition of greedy bisimulation, the carrier of \({≡}\) is a subset of the carrier of \({≊}\). 
    Then because of~\Cref{thm:greedy-bisim-iff-norm-bisim}, \(\norm(≊)\) is a bisimulation between \(\norm(S)\) and \(\norm(U)\), which implies \(\norm(≊) ⊑ {≡}\).
    Finally, because \(\norm\) do not change the carrier set, therefore the carrier of \(≡\) and \(≊\) necessarily coincide.
\end{proofEnd}

@@

\begin{theoremEnd}{theorem}[Correctness]\label{thm:greedy-bisim-correctness}
    Given two states in GKAT coalgebras \(s ∈ S\) and \(u ∈ ,U\), there exists a greedy bisimulation \({≈} ⊆ S × U\) s.t. \(s ≈ u\), if and only if \(⟦s⟧_S = ⟦u⟧_U\).
\end{theoremEnd}

@@

Before we introduce the algorithm, we will first briefly sketch the liveness-check algorithm. It uses depth-first search to check whether there exists any accepting states in the reachable states of \(s ∈ S\): if there exists an accepting state in \(⟨s⟩_S\), then the algorithm terminates immediately, and returns that \(s\) is live, otherwise it would return all the states in \(⟨s⟩\), and by~\cref{thm:dead-iff-all-reachable-dead}, all the states in \(⟨s⟩\) are dead. Note that we use depth-first search here for its simplicity, other search algorithms will also be sound.
